---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# smarterapi

<!-- badges: start -->
<!-- badges: end -->

The goal of `smarterapi` is to collect data from *SMARTER REST API*

## Installation

You can install the development version of smarterapi from [GitHub](https://github.com/) with:

```r
# install.packages("devtools")
devtools::install_github("cnr-ibba/r-smarter-api")
```

## SMARTER credentials

The *SMARTER REST API* is not yet public and can be accessed providing a valid
*token* in every *HTTP* request. `smarterapi` can deal with token generation and
management and implement functions in order to automatically manage tokens in 
every user requests. The function `get_smarter_token` can be used to manage tokens
and cache them inside `smarterapi` package. You could provide manually your 
SMARTER credentials by calling `get_smarter_token` with the proper parameters
in order to generate a token, however the recommended way is to store your
credentials using environment variables, and let `smarterapi` deal with 
your token management. Please remember that SMARTER credentials **MUST NOT be
tracked** in your code nor provided to un-authorized people. To manage your
environment properly, you should define such environment variables in your `$HOME`
directory, outside any other *R projects*. If you use *rstudio*, you can call
an utility function which open such environment file:

```r
usethis::edit_r_environ()
```

Variables in environment file should be defined with `key=value` syntax. In order
to configure properly `smarterapi`, you have to define `SMARTER_API_USERNAME` and 
`SMARTER_API_PASSWORD` with your SMARTER credentials, like this:

```text
SMARTER_API_USERNAME=<smarter username>
SMARTER_API_PASSWORD=<smarter password>
```

## Example

This is a basic example which shows you how to collect data from SMARTER REST API
for the *Italian* goats belonging to the *Adaptmap* dataset:

```{r example}
library(smarterapi)
datasets <- get_smarter_datasets(
  query = list(species = "Goat", search = "adaptmap", type = "genotypes"))
adatpmap_id <- datasets["_id.$oid"][1]
adaptmap_goats <- get_smarter_samples(
  species = "Goat", query = list(dataset=adatpmap_id, country="Italy"))
```
