---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# smarterapi

<!-- badges: start -->
[![R-CMD-check](https://github.com/cnr-ibba/r-smarter-api/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/cnr-ibba/r-smarter-api/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of `smarterapi` is to collect data from *SMARTER REST API*

## Installation

You can install the development version of smarterapi from [GitHub](https://github.com/) with:

```r
# install.packages("devtools")
devtools::install_github("cnr-ibba/r-smarter-api")
```

## SMARTER credentials

The *SMARTER REST API* is not yet public and can be accessed providing a valid
*token* in every *HTTP* request. `smarterapi` can deal with token generation and
management and implement functions in order to automatically manage tokens in 
every user requests. The function `get_smarter_token` can be used to manage tokens
and cache them inside `smarterapi` package. You could provide manually your 
SMARTER credentials by calling `get_smarter_token` with the proper parameters
in order to generate a token, however the recommended way is to store your
credentials using environment variables, and let `smarterapi` deal with 
your token management. Please remember that SMARTER credentials **MUST NOT be
tracked** in your code nor provided to un-authorized people. To manage your
environment properly, you should define such environment variables in your `$HOME`
directory, outside any other *R projects*. If you use *rstudio*, you can call
an utility function which open such environment file:

```r
usethis::edit_r_environ()
```

Variables in environment file should be defined with `key=value` syntax. In order
to configure properly `smarterapi`, you have to define `SMARTER_API_USERNAME` and 
`SMARTER_API_PASSWORD` with your SMARTER credentials, like this:

```text
SMARTER_API_USERNAME=<smarter username>
SMARTER_API_PASSWORD=<smarter password>
```

## Querying the SMARTER backend

`smarterapi` provides a set of functions used to fetch data from *SMARTER-backend*
endpoints described in [api documentation](https://webserver.ibba.cnr.it/smarter-api/docs/)
and returning them into a `data.frame` object. For example, the `get_smarter_datasets` 
lets to query the [Datasets endpoint](https://webserver.ibba.cnr.it/smarter-api/docs/#/Datasets/get_datasets),
while `get_smarter_samples` is able to query and parse the 
[Sample endpoints](https://webserver.ibba.cnr.it/smarter-api/docs/#/Samples)
response. Each `smarterapi` function is documented in `R` and you can get help on
each function like any other R functions. If you manage tokens using `smarterapi`
and environment variables, you don't need to provide a token as a parameter while
fetching data, those function will handle the token generation and authentication 
for you. There are two types of parameters that are required to fetch data through
the *SMARTER-backend*: the `species` parameter, which can be `Goat` or `Sheep` respectively
for goat and sheep [Samples](https://webserver.ibba.cnr.it/smarter-api/docs/#/Samples)
or [Variants](https://webserver.ibba.cnr.it/smarter-api/docs/#/Variants) endpoints,
and the `query` parameter which can be provided to any *get_smarter_* function.
The `species` parameter is *mandatory* in order to query an endpoint specific for
*Goat* or *Sheep*, while the `query` parameter is optional and need to be specified
as a `list()` object in order to limit your query to some data in particular. 
For example, if you need all the foreground genotypes datasets, you can collect
data like this:

```r
datasets <- get_smarter_datasets(query = list(type="genotypes", type="foreground"))
```

while if you require only background goat samples, you can do like this:

```r
goat_samples <- get_smarter_samples(species = "Goat", query=list(type="background"))
```

The full option list available to each endpoint is available on the SMARTER-backend
[swagger documentation](https://webserver.ibba.cnr.it/smarter-api/docs/): the option
name to use is the same name described in *parameters*, and description and parameter
types can give you hints on how to exploit endpoint properly. For instance, parameters
described as *array of objects* can be specified multiple times:

```r
> goat_breeds <- get_smarter_breeds(query = list(species="Goat", search="land"))
> goat_breeds[c("name","code")]
            name code
1      Rangeland  RAN
2       Landrace  LNR
3         Landin  LND
4 Icelandic goat  ICL
>
> goat_samples <- get_smarter_samples(
    species = "Goat", 
    query = list(
      breed_code="RAN", 
      breed_code="LNR", 
      breed_code="LND", 
      breed_code="ICL"
    )
  )
```

## Example

This is a basic example which shows you how to collect data from SMARTER REST API
for the *Italian* goats belonging to the *Adaptmap* dataset:

```{r example}
library(smarterapi)
datasets <- get_smarter_datasets(
  query = list(species = "Goat", search = "adaptmap", type = "genotypes", type="background"))
adatpmap_id <- datasets["_id.$oid"][1]
adaptmap_goats <- get_smarter_samples(
  species = "Goat", query = list(dataset=adatpmap_id, country="Italy"))
```
