---
title: "Citing SMARTER data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Citing SMARTER data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette explains how to properly cite SMARTER data when used in your
research work. SMARTER data are composed by *background* datasets and
*foreground* datasets: the former derive from other public sources, while the latter
are generated by the SMARTER project. Ideally, when using SMARTER data, *foreground*
datasets should be cited using the SMARTER-database publication (see below), while
*background* datasets should be cited using the original publication(s) from which
data were taken. In this vignette we provide some tips to retrieve citation
information using the `smarterapi` package.

## Loading the package

When loading the `smarterapi` package, a message is shown to remind users
to cite SMARTER data if used in publications:

```{r setup}
library(smarterapi)
```

You can also retrieve citation information using the `citation()` function:

```{r citation}
citation("smarterapi")
```

This will provide you with the reference to the SMARTER-database publication to
be used when citing SMARTER data, in particular for *foreground* datasets.
