---
title: "Working with GeoJSON data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with GeoJSON data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette explain some tips in order to work and visualize SMARTER data
using the [GeoJSON endpoints](https://webserver.ibba.cnr.it/smarter-api/docs/#/GeoJSON):
These endpoints in particular retrieve all the samples with GPS coordinates
(like using the `locations__exists=TRUE` parameter with the `get_smarter_samples()`
method) and then return results as *geo dataframe*. First, load some required
libraries:

```{r setup}
library(smarterapi)
library(sf)
```

Next collect data using the `smarterapi` methods. For example for the *Sheep
hapmap background* dataset:

```{r}
hapmap_dataset <- get_smarter_datasets(query = list(
  species = "Sheep",
  type = "background",
  type = "genotypes",
  search = "hapmap"
))
hapmap_data <- get_smarter_geojson(species = "Sheep", query = list(
  dataset = hapmap_dataset["_id.$oid"][1],
  country = "Italy"
))
```

The `get_smarter_geojson()` methods returns a `sf()` *MULTIPOINTS* objects since
multiple locations could be collected for each animal, for example for 
*transhumant* samples. However, for simplicity, we can focus our analyses using
only one GPS coordinate, transforming this *MULTIPOINTS* object into a *POINT*
object:

```{r warning=FALSE}
hapmap_data <- hapmap_data %>% sf::st_cast("POINT", do_split = FALSE)
hapmap_data
```

The *warnings* you see in your terminal tell you that you are considering only
the first coordinate object from each *MULTIPOINT* collection.
